---
title: "The Story: How the Three Took Over"
subtitle: "A scrollytelling narrative centered on Stephen Curry's influence"
format:
  html:
    theme: darkly
    toc: false
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
source("R/prepare_data.R")

nba <- load_shot_data(sample_per_group = 1200L)
comp <- build_player_comparison_data(min_career_fga = 1000L, sample_made_per_player_season = 30L)

season_summ <- nba$season_summ
selection_long <- nba$selection_long

season_breaks <- season_summ$season
season_label_lookup <- setNames(season_summ$season_label, season_summ$season)
label_season <- function(x) unname(season_label_lookup[as.character(x)])

steph_year <- comp$player_year |>
  filter(player_name == "Stephen Curry") |>
  arrange(season)

peak_row <- steph_year |>
  slice_max(fga3, n = 1, with_ties = FALSE)
peak_season <- peak_row$season[[1]]
peak_season_label <- peak_row$season_label[[1]]
peak_3pa <- peak_row$fga3[[1]]
peak_3pm <- peak_row$fgm3[[1]]

peer_trends <- comp$player_year |>
  filter(player_name %in% c("Stephen Curry", "Damian Lillard", "James Harden", "Klay Thompson"))

impact <- comp$league_year |>
  left_join(
    steph_year |>
      transmute(season, steph_tpa_per_game = tpa_per_game),
    by = "season"
  ) |>
  mutate(
    league_three_rate_index = 100 * league_three_rate / first(league_three_rate),
    steph_tpa_index = 100 * steph_tpa_per_game / first(steph_tpa_per_game)
  )

injury_season <- 2019L
injury_label <- season_label_from_start(injury_season)
covid_season <- injury_season
covid_label <- injury_label
```

::: {.story-lead}
People attempted threes before Steph Curry. The shift is that after his peak stretch, the league restructured around that shot profile.  
Even in **`r injury_label`**, when Curry was mostly absent, league three-point behavior stayed elevated.
:::

::: {.callout-tip}
## Presentation Context
The project deck is available here: [INFO 3312 Slides (PDF)](assets/slides/INFO-3312-Slides.pdf){target="_blank"}.

Narrative anchor from the presentation: league 3PA share rose from about **22% (2004-05)** to over **42% (2023-24)** while midrange share dropped from about **30%** to under **12%**.
:::

::: {.callout-note}
## Injury Context Used In The Visuals
In the **2019-20** season, Stephen Curry broke his left hand (second metacarpal) on **October 30, 2019**, needed surgery, and missed **58 straight games**.  
He played only **5 games** that season (also affected by the COVID suspension), and Golden State finished **15-50**.  
He briefly returned in March 2020, then used the long break to rehab into a more efficient post-injury version.
:::

::: {.story-scrolly}
::: {.story-graphic}

::: {.story-figure data-figure="1"}
```{r}
ggplot(season_summ, aes(season, three_rate)) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = min(season_summ$season) - 0.5,
    xmax = 2014.5,
    ymin = -Inf,
    ymax = Inf,
    fill = "#4a5568",
    alpha = 0.22
  ) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = injury_season - 0.5,
    xmax = injury_season + 0.5,
    ymin = -Inf,
    ymax = Inf,
    fill = "#f94144",
    alpha = 0.14
  ) +
  geom_line(linewidth = 1.4, color = "#8ecae6") +
  geom_point(size = 2.1, color = "#8ecae6") +
  annotate(
    "text",
    x = injury_season,
    y = max(season_summ$three_rate, na.rm = TRUE) * 0.96,
    label = paste0("Curry injury + COVID: ", injury_label),
    size = 3.5,
    color = "#fbc4ab"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = season_breaks, labels = label_season(season_breaks)) +
  labs(
    title = "Before the inflection",
    subtitle = "League 3PA share was growing, but slowly",
    x = NULL,
    y = "3PA share"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```
:::

::: {.story-figure data-figure="2"}
```{r}
peer_breaks <- sort(unique(peer_trends$season))

ggplot(peer_trends, aes(season, tpa_per_game, color = player_name)) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = injury_season - 0.5,
    xmax = injury_season + 0.5,
    ymin = -Inf,
    ymax = Inf,
    fill = "#f94144",
    alpha = 0.12
  ) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 1.6) +
  geom_vline(xintercept = injury_season, linetype = "dashed", color = "#fbc4ab") +
  scale_x_continuous(breaks = peer_breaks, labels = label_season(peer_breaks)) +
  labs(
    title = "Not just threes, but volume at scale",
    subtitle = "Curry separates from elite peers in sustained 3PA/game, including the injury window",
    x = NULL,
    y = "3PA per game",
    color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```
:::

::: {.story-figure data-figure="3"}
```{r}
steph_peak_plot <- steph_year |>
  mutate(peak_flag = if_else(season == peak_season, "Peak season", "Other seasons"))

ggplot(steph_peak_plot, aes(season, fga3, fill = peak_flag)) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = injury_season - 0.5,
    xmax = injury_season + 0.5,
    ymin = -Inf,
    ymax = Inf,
    fill = "#f94144",
    alpha = 0.12
  ) +
  geom_col(width = 0.72) +
  geom_text(
    data = steph_peak_plot |> filter(season == peak_season),
    aes(label = paste0("Peak: ", comma(fga3), " attempts\n", comma(fgm3), " makes")),
    vjust = -0.35,
    size = 4.1,
    color = "#111"
  ) +
  scale_fill_manual(values = c("Peak season" = "#ffb703", "Other seasons" = "#6aa5d8")) +
  scale_x_continuous(breaks = steph_peak_plot$season, labels = steph_peak_plot$season_label) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Curry's peak season set the benchmark",
    subtitle = paste0("Most 3PA in one season: ", peak_season_label, " | Injury break highlighted at ", injury_label),
    x = NULL,
    y = "3PA",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```
:::

::: {.story-figure data-figure="4"}
```{r}
impact_long <- bind_rows(
  impact |>
    transmute(season, series = "League 3PA rate index", value = league_three_rate_index),
  impact |>
    transmute(season, series = "Steph 3PA/game index", value = steph_tpa_index)
)

impact_breaks <- impact$season

ggplot(impact_long, aes(season, value, color = series)) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = injury_season - 0.5,
    xmax = injury_season + 0.5,
    ymin = -Inf,
    ymax = Inf,
    fill = "#f94144",
    alpha = 0.12
  ) +
  geom_line(linewidth = 1.35) +
  geom_point(size = 1.7) +
  geom_vline(xintercept = injury_season, linetype = "dashed", color = "#fbc4ab") +
  geom_vline(xintercept = c(2014, 2016), linetype = "dashed", color = "#9aa1aa") +
  scale_x_continuous(breaks = impact_breaks, labels = label_season(impact_breaks)) +
  labs(
    title = "2014-2016 is the acceleration window",
    subtitle = "League behavior catches up after Curry's explosive years",
    x = NULL,
    y = "Index (2013-14 = 100)",
    color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```
:::

::: {.story-figure data-figure="5"}
```{r}
ggplot(selection_long, aes(season, share, fill = bucket)) +
  geom_area(color = "#0d1117", linewidth = 0.2) +
  geom_vline(xintercept = injury_season, color = "#fbc4ab", linetype = "dotted") +
  annotate(
    "text",
    x = injury_season,
    y = 0.98,
    label = paste0("Injury season: ", injury_label),
    size = 3.3,
    color = "#fbc4ab"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = season_breaks, labels = label_season(season_breaks)) +
  scale_fill_manual(
    values = c(
      "Restricted Area" = "#8ecae6",
      "Paint (non-RA)" = "#7db9de",
      "Midrange" = "#adb5bd",
      "Corner 3" = "#f4a261",
      "Above the Break" = "#e76f51"
    )
  ) +
  labs(
    title = "The modern shot diet",
    subtitle = "Above-the-break threes scale up while midrange compresses",
    x = NULL,
    y = "Share of attempts",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```
:::

:::

::: {.story-steps}

::: {.story-step .align-right data-figure="1"}
::: {.step-card}
### 1) Baseline era: before the inflection
Threes were always part of NBA offense, but the league mix was still more traditional.  
This section sets up the "before" frame from the project objective.
:::
:::

::: {.story-step .align-left data-figure="2"}
::: {.step-card}
### 2) Curry's volume ceiling is the mechanism
Curry did not only shoot efficiently; he normalized high-volume pull-up threes over full seasons.
:::
:::

::: {.story-step .align-center data-figure="3"}
::: {.step-card}
### 3) Peak year analysis
His highest-volume three-point season reached **`r comma(peak_3pa)` attempts** and **`r comma(peak_3pm)` makes** in **`r peak_season_label`**.  
This is the point where "influence" becomes measurable against league trend movement.
:::
:::

::: {.story-step .align-right data-figure="4"}
::: {.step-card}
### 4) 2014-2016 hypothesis window
This is where Curry's rise and the league's acceleration line up most clearly, matching your core hypothesis.
:::
:::

::: {.story-step .align-left data-figure="5"}
::: {.step-card}
### 5) Structural shift, then COVID disruption
By **`r injury_label`**, Curry was mostly out, but league three-point volume still held.  
That supports the story conclusion: the shot diet change persisted beyond one player being active game-to-game.
:::
:::

:::
:::

<script>
document.addEventListener("DOMContentLoaded", function () {
  const steps = Array.from(document.querySelectorAll(".story-step"));
  const figures = Array.from(document.querySelectorAll(".story-figure"));

  const activateFigure = (figureId) => {
    figures.forEach((figure) => {
      const isActive = figure.dataset.figure === figureId;
      figure.classList.toggle("is-active", isActive);
    });
    steps.forEach((step) => {
      const isActive = step.dataset.figure === figureId;
      step.classList.toggle("is-active", isActive);
    });
  };

  const updateActiveFromViewport = () => {
    const focusY = window.innerHeight * 0.48;
    let bestStep = null;
    let bestDistance = Infinity;

    steps.forEach((step) => {
      const rect = step.getBoundingClientRect();
      const inView = rect.bottom > 0 && rect.top < window.innerHeight;
      if (!inView) return;

      const centerY = rect.top + rect.height / 2;
      const distance = Math.abs(centerY - focusY);
      if (distance < bestDistance) {
        bestDistance = distance;
        bestStep = step;
      }
    });

    if (bestStep) activateFigure(bestStep.dataset.figure);
  };

  let ticking = false;
  const onScroll = () => {
    if (ticking) return;
    ticking = true;
    requestAnimationFrame(() => {
      updateActiveFromViewport();
      ticking = false;
    });
  };

  window.addEventListener("scroll", onScroll, { passive: true });
  window.addEventListener("resize", updateActiveFromViewport);
  updateActiveFromViewport();
});
</script>
