---
title: "Steph vs Player"
format:
  html:
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
source("R/prepare_data.R")

dir.create("docs/data", recursive = TRUE, showWarnings = FALSE)
comp <- build_player_comparison_data(min_career_fga = 1000L, sample_made_per_player_season = 30L)

write_csv(comp$player_year, "docs/data/player_year.csv")
write_csv(comp$made_shots, "docs/data/made_shots.csv")
write_csv(comp$steph_league, "docs/data/steph_league.csv")
```

This view tests the hypothesis that **Stephen Curry's 2014-2016 stretch** accelerated league three-point behavior.

::: {.callout-note}
## Injury Note For Interpretation
Stephen Curry broke his left hand on **October 30, 2019** (season **2019-20**), underwent surgery, and missed **58 consecutive games**.  
He played only **5 games** that season, briefly returned in March 2020, and Golden State finished **15-50**.  
Charts below mark **2019-20** to show the break in form and the league's continued three-point behavior.
:::

```{r}
library(ggplot2)

steph_2014_2016 <- comp$made_shots |>
  filter(player_name == "Stephen Curry", season >= 2014, season <= 2016) |>
  mutate(loc_y_flipped = 47 - loc_y)

if (requireNamespace("sportyR", quietly = TRUE)) {
  sportyR::geom_basketball(league = "NBA", display_range = "offense", rotation = 90) +
    geom_point(
      data = steph_2014_2016,
      aes(x = loc_x, y = loc_y_flipped),
      color = "#1f77b4",
      alpha = 0.2,
      size = 0.7
    ) +
    coord_cartesian(xlim = c(-25, 25), ylim = c(0, 47)) +
    labs(title = "sportyR half-court: Stephen Curry made shots (2014-15 to 2016-17)") +
    theme_void()
} else {
  ggplot(steph_2014_2016, aes(x = loc_x, y = loc_y_flipped)) +
    geom_point(color = "#1f77b4", alpha = 0.2, size = 0.7) +
    coord_equal(xlim = c(-25, 25), ylim = c(0, 47), expand = FALSE) +
    labs(
      title = "Half-court made shots (fallback render)",
      subtitle = "sportyR unavailable in this environment",
      x = NULL,
      y = NULL
    ) +
    theme_void()
}
```

```{ojs}
playerYear = await d3.csv("data/player_year.csv", d3.autoType)
madeShots = await d3.csv("data/made_shots.csv", d3.autoType)
stephLeague = await d3.csv("data/steph_league.csv", d3.autoType)

players = Array.from(new Set(playerYear.map(d => d.player_name))).sort((a, b) => d3.ascending(a, b))
compareOptions = players.filter(p => p !== "Stephen Curry")

seasonMeta = Array.from(
  new Set(playerYear.map(d => +d.season))
).sort((a, b) => a - b).map(season => ({
  season,
  season_label: `${season}-${String((season + 1) % 100).padStart(2, "0")}`
}))

seasonLookup = new Map(seasonMeta.map(d => [d.season, d.season_label]))
seasonFromLabel = new Map(seasonMeta.map(d => [d.season_label, d.season]))
seasonLabels = seasonMeta.map(d => d.season_label)
injurySeason = 2019
injuryLabel = seasonLookup.get(injurySeason) ?? "2019-20"
injuryNote = `Injury context: broken left hand (${injuryLabel}), 58-game absence, 5 games played, Warriors 15-50.`

typeaheadInput = (options, {label, value, placeholder, maxSuggestions = 5}) => {
  const listId = `suggest-${Math.random().toString(36).slice(2, 10)}`
  const wrapper = html`<label style="display:grid; gap:0.35rem;">
    <span>${label} (search)</span>
    <input type="text" list="${listId}" value="${value ?? ""}" placeholder="${placeholder ?? ""}" style="width:100%;">
    <datalist id="${listId}"></datalist>
    <small style="opacity:0.75;">Top ${maxSuggestions} suggestions while typing</small>
  </label>`
  const input = wrapper.querySelector("input")
  const datalist = wrapper.querySelector("datalist")
  const optionStrings = options.map(option => String(option))

  const refreshSuggestions = () => {
    const query = input.value.trim().toLowerCase()
    const matches = optionStrings
      .filter(option => option.toLowerCase().includes(query))
      .slice(0, maxSuggestions)
    datalist.replaceChildren(...matches.map(match => html`<option value="${match}"></option>`))
  }

  const emitInput = () => {
    wrapper.value = input.value
    wrapper.dispatchEvent(new Event("input", {bubbles: true}))
  }

  input.addEventListener("input", () => {
    refreshSuggestions()
    emitInput()
  })
  input.addEventListener("change", emitInput)

  wrapper.value = input.value
  refreshSuggestions()
  return wrapper
}

adaptiveInput = (options, config) => (
  options.length > 10
    ? typeaheadInput(options, config)
    : Inputs.select(options, {label: config.label, value: config.value})
)

resolveSearchValue = (inputValue, options, fallback) => {
  if (options.length === 0) return fallback ?? null
  const fallbackValue = options.includes(fallback) ? fallback : options[0]
  const query = String(inputValue ?? "").trim().toLowerCase()
  if (!query) return fallbackValue
  const exact = options.find(option => String(option).toLowerCase() === query)
  if (exact) return exact
  const partial = options.find(option => String(option).toLowerCase().includes(query))
  return partial ?? fallbackValue
}

defaultStartLabel = seasonLookup.get(2014) ?? seasonLabels[0]
defaultEndLabel = seasonLookup.get(2016) ?? seasonLabels[seasonLabels.length - 1]

viewof comparePlayerInput = adaptiveInput(compareOptions, {
  label: "Compare player",
  value: "Damian Lillard",
  placeholder: "Type player name..."
})

comparePlayer = resolveSearchValue(comparePlayerInput, compareOptions, "Damian Lillard")

viewof seasonStartInput = adaptiveInput(seasonLabels, {
  label: "Season start",
  value: defaultStartLabel,
  placeholder: "Type season, e.g. 2019-20"
})

viewof seasonEndInput = adaptiveInput(seasonLabels, {
  label: "Season end",
  value: defaultEndLabel,
  placeholder: "Type season, e.g. 2020-21"
})

seasonStartLabel = resolveSearchValue(seasonStartInput, seasonLabels, defaultStartLabel)
seasonEndLabel = resolveSearchValue(seasonEndInput, seasonLabels, defaultEndLabel)
seasonStart = seasonFromLabel.get(seasonStartLabel)
seasonEnd = seasonFromLabel.get(seasonEndLabel)
startSeason = Math.min(seasonStart, seasonEnd)
endSeason = Math.max(seasonStart, seasonEnd)
windowLabel = `${seasonLookup.get(startSeason)} to ${seasonLookup.get(endSeason)}`

windowTeamOptions = ["All", ...Array.from(new Set(
  madeShots
    .filter(d => d.season >= startSeason && d.season <= endSeason)
    .map(d => d.team_name)
    .filter(Boolean)
)).sort((a, b) => d3.ascending(a, b))]

viewof teamFilterInput = adaptiveInput(windowTeamOptions, {
  label: "Team filter",
  value: "All",
  placeholder: "Type team name..."
})

teamFilter = resolveSearchValue(teamFilterInput, windowTeamOptions, "All")

inWindow = d => d.season >= startSeason && d.season <= endSeason
withTeam = d => teamFilter === "All" || d.team_name === teamFilter

stephRows = playerYear.filter(d => d.player_name === "Stephen Curry" && inWindow(d) && withTeam(d))
compRows = playerYear.filter(d => d.player_name === comparePlayer && inWindow(d) && withTeam(d))

summarizePlayer = rows => {
  const fga = d3.sum(rows, d => d.fga)
  const fgm = d3.sum(rows, d => d.fgm)
  const fga3 = d3.sum(rows, d => d.fga3)
  const fgm3 = d3.sum(rows, d => d.fgm3)
  const games = d3.sum(rows, d => d.games)
  return {
    seasons: rows.length,
    fg_pct: fga > 0 ? fgm / fga : NaN,
    tp_pct: fga3 > 0 ? fgm3 / fga3 : NaN,
    efg: fga > 0 ? (fgm + 0.5 * fgm3) / fga : NaN,
    tpa_per_game: games > 0 ? fga3 / games : NaN,
    tpa_per_season: rows.length > 0 ? fga3 / rows.length : NaN
  }
}

stephAgg = summarizePlayer(stephRows)
compAgg = summarizePlayer(compRows)

fmtPct = v => Number.isFinite(v) ? d3.format(".1%")(v) : "n/a"
fmtNum = v => Number.isFinite(v) ? d3.format(".2f")(v) : "n/a"

md`
### Window summary (${windowLabel}) ${teamFilter === "All" ? "" : `| Team: ${teamFilter}`}

${injuryNote}

| Metric | Stephen Curry | ${comparePlayer} |
|---|---:|---:|
| FG% | ${fmtPct(stephAgg.fg_pct)} | ${fmtPct(compAgg.fg_pct)} |
| 3P% | ${fmtPct(stephAgg.tp_pct)} | ${fmtPct(compAgg.tp_pct)} |
| eFG% | ${fmtPct(stephAgg.efg)} | ${fmtPct(compAgg.efg)} |
| 3PA / game | ${fmtNum(stephAgg.tpa_per_game)} | ${fmtNum(compAgg.tpa_per_game)} |
| 3PA / season | ${fmtNum(stephAgg.tpa_per_season)} | ${fmtNum(compAgg.tpa_per_season)} |
`
```

```{ojs}
trendRows = playerYear
  .filter(d => (d.player_name === "Stephen Curry" || d.player_name === comparePlayer))
  .filter(d => teamFilter === "All" || d.team_name === teamFilter)
  .map(d => ({...d, player: d.player_name}))

Plot.plot({
  width: 930,
  height: 380,
  x: {label: "Season", tickFormat: d => seasonLookup.get(d) ?? String(d)},
  y: {label: "3PA per game"},
  color: {legend: true},
  marks: [
    Plot.ruleX([injurySeason], {stroke: "#d00000", strokeDasharray: "4,4"}),
    Plot.lineY(trendRows, {x: "season", y: "tpa_per_game", stroke: "player"}),
    Plot.dot(trendRows, {x: "season", y: "tpa_per_game", fill: "player"})
  ]
})
```

```{ojs}
Plot.plot({
  width: 930,
  height: 380,
  x: {label: "Season", tickFormat: d => seasonLookup.get(d) ?? String(d)},
  y: {label: "3P%", percent: true},
  color: {legend: true},
  marks: [
    Plot.ruleX([injurySeason], {stroke: "#d00000", strokeDasharray: "4,4"}),
    Plot.lineY(trendRows, {x: "season", y: "tp_pct", stroke: "player"}),
    Plot.dot(trendRows, {x: "season", y: "tp_pct", fill: "player"})
  ]
})
```

```{ojs}
arcPoints = (cx, cy, r, startDeg, endDeg, steps = 80) => {
  const pts = []
  const start = startDeg * Math.PI / 180
  const end = endDeg * Math.PI / 180
  for (let i = 0; i <= steps; i++) {
    const t = start + (end - start) * (i / steps)
    pts.push({x: cx + r * Math.cos(t), y: cy + r * Math.sin(t)})
  }
  return pts
}

courtSegments = () => {
  const segs = []
  const pushLine = (a, b, id) => segs.push({id, points: [{x: a[0], y: a[1]}, {x: b[0], y: b[1]}]})

  pushLine([-25, 0], [25, 0], "baseline")
  pushLine([-25, 0], [-25, 47], "left_sideline")
  pushLine([25, 0], [25, 47], "right_sideline")
  pushLine([-25, 47], [25, 47], "halfcourt")
  pushLine([-8, 0], [-8, 19], "paint_left")
  pushLine([8, 0], [8, 19], "paint_right")
  pushLine([-8, 19], [8, 19], "paint_top")

  segs.push({id: "restricted", points: arcPoints(0, 5.25, 4, 0, 180, 70)})
  segs.push({id: "freethrow", points: arcPoints(0, 19, 6, 0, 360, 120)})
  pushLine([-22, 0], [-22, 14], "corner3_left")
  pushLine([22, 0], [22, 14], "corner3_right")
  segs.push({id: "arc3", points: arcPoints(0, 5.25, 23.75, 22, 158, 120)})

  return segs
}

segmentRows = courtSegments().flatMap(seg =>
  seg.points.map((p, i) => ({id: seg.id, order: i, x: p.x, y: p.y}))
)

windowShotsBase = madeShots.filter(d => d.season >= startSeason && d.season <= endSeason)
windowShotsFiltered = teamFilter === "All" ? windowShotsBase : windowShotsBase.filter(d => d.team_name === teamFilter)
stephShots = windowShotsFiltered.filter(d => d.player_name === "Stephen Curry")
compShots = windowShotsFiltered.filter(d => d.player_name === comparePlayer)

shotMap = (rows, title, color) => Plot.plot({
  width: 450,
  height: 430,
  title,
  subtitle: injuryNote,
  margin: 18,
  x: {domain: [-25, 25], axis: null},
  y: {domain: [0, 47], axis: null},
  marks: [
    Plot.line(segmentRows, {x: "x", y: "y", z: "id", stroke: "#666", strokeWidth: 1}),
    Plot.dot(rows, {x: "loc_x", y: "loc_y", r: 1.5, fill: color, opacity: 0.33, tip: true})
  ]
})

html`<div style="display:flex;gap:20px;flex-wrap:wrap;">
  <div>${shotMap(stephShots, `Stephen Curry made shots (${windowLabel})`, "#1f77b4")}</div>
  <div>${shotMap(compShots, `${comparePlayer} made shots (${windowLabel})`, "#d95f02")}</div>
</div>`
```

```{ojs}
impactRows = stephLeague.map(d => ({
  season: d.season,
  series: "League 3PA rate index",
  value: d.league_three_rate_index
})).concat(
  stephLeague.map(d => ({
    season: d.season,
    series: "Steph 3PA/G index",
    value: d.steph_tpa_index
  }))
)

Plot.plot({
  width: 930,
  height: 360,
  subtitle: injuryNote,
  x: {label: "Season", tickFormat: d => seasonLookup.get(d) ?? String(d)},
  y: {label: "Index (2013-14 = 100)"},
  color: {legend: true},
  marks: [
    Plot.ruleX([2019], {stroke: "#d00000", strokeDasharray: "3,3"}),
    Plot.lineY(impactRows, {x: "season", y: "value", stroke: "series"}),
    Plot.dot(impactRows, {x: "season", y: "value", fill: "series"}),
    Plot.ruleX([2014, 2016], {stroke: "#999", strokeDasharray: "4,4"})
  ]
})
```
